---
title: "Digital Humanities Intro to R"
author: "Gregory J. Matthews"
date: "2023-11-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Contents 
 - Basic of R
 - Data Structures
 - Input and output
 - Tidyverse
 - Data Viz
 - Clustering Example (time permitting)

## Intro to R and R Studio 
 - R is a language
 - R Studio is an integrated development environment (IDE)

## Basics of R

```{r}
####################################
#Basics of R
####################################

#Big note!
# R starts indexing at 1 NOT ZERO!!!!!

####################################
#Assignment
####################################
#Preferred syntax
height <- 442

#These also work
442 -> height
height = 442

print(height)
#don't need print in R
height

####################################
#Comments
####################################
#This is how you make a comment!

####################################
#Variables
####################################
#Default to numeric
height <- 442 
class(height)

height <- 442.0
class(height)

#Character
height <- "really tall"
class(height)

#Can force it to be an integer
height <- as.integer(442)
class(height)
object.size(height)

####################################
#R is case sensitive
####################################
#These are all different objects
name <- "greg" 
Name <- "steve"
NAME <- "mary"

####################################
#Conditional Statements
####################################
#If
sky <- "sunny"
if (sky == "sunny"){
  print("Leave your umbrella at home!")
}

sky <- "cloudy"
if (sky == "sunny"){
  print("Leave your umbrella at home!")
}

#Ifelse

sky <- "cellphone"
if (sky == "sunny"){
  print("Leave your umbrella at home!")
} else {
  print("Bring your umbrella")
}

#Nested
sky <- "sunny"
if (sky == "sunny") {
  print("Leave your umbrella at home!")
} else {
  if (sky == "cloudy") {
    print("Bring your umbrella")
  } else {
    if (sky == "snowing") {
      print("Grab your parka")
    } else {
      print("Your guess is as good as mine")
    }
  }
}

#But we can use elseif 
sky <- "dsfsad"
if (sky == "sunny") {
  print("Leave your umbrella at home!")
} else if (sky == "cloudy") {
  print("Bring your umbrella")
} else if (sky == "snowing") {
  print("Grab your parka")
} else {
  print("Your guess is as good as mine")
}
  
#There is also the function ifelse
sky <- "sunny"
#ifelse(test,yes,no)
ifelse(sky == "sunny","Sunny","not sunny")
#No print statement needed!

####################################
#Conditional Statements
####################################
#Booleans
#TRUE and FALSE
#Need to be all capital letters!
a <- TRUE
b <- FALSE
#TRUE is a 1; FALSE is a 0
4 + a

d <- 0
if (d == FALSE){
  print("d is FALSE")
}

#also works
d <- TRUE
if (d){
  print("d is TRUE")
}


#Also works
d <- 1
if (d){
  print("d is TRUE")
}

#Integers
a <- as.integer(37)
class(a)

#Floating Points / Numeric
b <- 37
class(b)

#Complex numbers
#We won't really use these this semester.  Or ever?
z <- 1 + 2i
class(z)


#Operations
#Add
3 + 3
#Subtract
6 - 3
#multiply
4 * 3
#divide
6 / 2
#floored quotient
floor(7 / 2)
#ceiling quotient
ceiling(7 / 2)
#remainder or modulus
7 %% 2
#absolute value
abs(-5)
#Complex numbers
4 + 7i
class(4 + 7i)
#Complex(length,real,imaginary)
#This will recycle as needed
complex(3,4,7:5)
#exponents
3^4
3**4

#Convert to different types
#Convert numeric to integer
a <- 5
class(a)
a
a <- as.integer(a)
class(a)

#Convert character to numberic
#Works if the character is a number
a <- "5"
class(a)
a <- as.numeric(a)
a
class(a)

#Convert a character to a numeric
#This doesn't work
a <- "five"
class(a)
a <- as.numeric(a)
a
class(a)

#Convert a numeric to a character
a <- 5
class(a)
a <- as.character(a)
a
class(a)



#Comparisons
# > Greater than
# < Less than
# >= Greater than or equal
# <= Less than or equal
# != Not equal

a <- 1
b <- 2
c <- 3
if (b >= a & b <= c){
  print("b is between a and c")
}

a <- 1
b <- 2
c <- 3
if (!(b < a | b > c)){
  print("b is between a and c")
}

#floating points numbers are inexact decimal values 
a <- 2.1 + 4.2
a == 6.3
a
class(a)

#Also numeric
a <- 3e3
class(a)

#Operations
a <- 1/2 
b <- 10
#Square root
sqrt(a)
#Trig
sin(a)
cos(a)
tan(a)
#Log base e
log(a)
#log base b
log(a, b)
#exponentiate
exp(a)

#Functions
#This function will return the larger of the two input numbers
find_max <- function(a, b) {
  if (a > b) {
    return(a)
  } else {
    return(b)
  }
}

find_max(1,2)
find_max(5,3)
```

## Data Structures 
 - Vectors
 - Matrices
 - Lists
 - ***Data Frames***

```{r}
#Vectors 
#One dimensional
vec <- c(7,8,3,4,10)

#Matrices
mat <- matrix(c(1,2,3,4), nrow = 2, ncol = 2)

#list
lll <- list(x = 3, y = vec, z = mat)
```

## Data Frames
```{r}
df <- data.frame(person = c("Greg","Steve","Mary"), 
           height = c(67, 68, 60),
           age = c(41, 20, 42), 
           inschool = c(FALSE,TRUE, TRUE))
```




## XML 

```{r}
#https://www.geeksforgeeks.org/working-with-xml-files-in-r-programming/
library("XML")
library("methods")
 
# the contents of sample.xml are parsed
data <- xmlParse(file = "/Users/gregorymatthews/Dropbox/DSCI401/sample.xml")
 
print(data)
```

```{r}
# Give the input file name to the function.
res <- xmlParse(file = "/Users/gregorymatthews/Dropbox/DSCI401/sample.xml")
 
# Extract the root node.
rootnode <- xmlRoot(res)

# number of nodes in the root.
nodes <- xmlSize(rootnode)
 
# get entire contents of a record
second_node <- rootnode[2]
 
# get 3rd attribute of 4th record
attr <- rootnode[[4]][[3]]
 
cat('number of nodes: ', nodes)
print ('details of 2 record: ')
print (second_node)
 
# prints the marks of the fourth record
print ('3rd attribute of 4th record: ', attr)

```


```{r}
#Convert to data frame.
dataframe <- xmlToDataFrame( "/Users/gregorymatthews/Dropbox/DSCI401/sample.xml")
print(dataframe)
```

## Input and output
```{r}
CIACountries <- read.csv("https://raw.githubusercontent.com/gjm112/DSCI401/main/data/CIACountries.csv")

#Output
#write.csv()
```

## Tidyverse
# Tidyverse grammar
- select(): Returns a sub set of columns (i.e. features or variables)
- filter(): Returns a subset of rows (i.e. observations)
- mutate(): add or modify existing columns
- arrange(): sort the rows
- summarize(): aggregate the data across rows (e.g., group it according to some criteria)

```{r}
CIACountries %>% filter(gdp > 10000) %>% select(pop, area, oil_prod)
#Doesn't work!  Order matters!
#CIACountries %>% select(pop, area, oil_prod)  %>% filter(gdp > 10000) 

CIACountries %>% mutate(density = pop/area)

CIACountries %>% arrange(-pop) %>% head()

CIACountries %>% group_by(net_users) %>% summarize(total_pop = sum(pop), n = n())


```


#Filter and Select
```{r}
library(tidyverse)
library(mdsr)
#Subset of rows
filter(CIACountries, pop > 1000000000)

#I can also do it like this
CIACountries %>% filter(pop > 1e9)

#In base R
CIACountries[CIACountries$pop > 1e9,]

#or
subset(CIACountries, pop > 1e9)

```


```{r}
#Subset of columns
select(CIACountries, country, pop, gdp)

#I can also do it like this
CIACountries %>% select(country, pop, gdp)

#In base R
CIACountries[, c("country", "pop", "gdp")]

#or
subset(CIACountries, select = c("country", "pop", "gdp"))

```


```{r}
#filter and select
filter(select(CIACountries, country, pop, gdp), pop > 1e9)

#Or we can do this with a pipeline
CIACountries %>% select(country, pop, gdp) %>% filter(pop > 1e9)

#Or in base R
CIACountries[CIACountries$pop > 1e9,c("country", "pop", "gdp")]

#or this:
subset(CIACountries, pop > 1e9, select = c("country", "pop", "gdp"))
```
#Mutate and Rename
```{r}
#Mutate
#Population Density
my_CIACountries <- mutate(CIACountries, dens = pop/area)
my_CIACountries[1,]

my_CIACountries <- CIACountries %>% mutate(dens = pop/area)
my_CIACountries[1,]

#Base R
library(mdsr)
my_CIACountries$dens <- CIACountries$pop / CIACountries$area

#rename
my_CIACountries <- CIACountries %>% rename(education = educ)
my_CIACountries[1,]
```

#arrange

```{r}
#Sorting with arrange
#Defaults to ascending
CIACountries_sort <- arrange(CIACountries, gdp)
CIACountries_sort[1:5,]

#Sort descending
CIACountries_sort <- arrange(CIACountries, desc(gdp))
CIACountries_sort[1:5,]

CIACountries_sort <- CIACountries %>% arrange(desc(gdp))
CIACountries_sort[1:5,]

#Base R with order
CIACountries_sort <- CIACountries[order(CIACountries$gdp),]
CIACountries_sort[1:5,]

CIACountries_sort <- CIACountries[order(-CIACountries$gdp),]
CIACountries_sort[1:5,]
```

#Summarize and Group_by
```{r}
#What is the average area by high vs low education countries?
my_CIACountries <-
  CIACountries %>% mutate(high_educ = educ > 4.5) %>% group_by(high_educ) %>% summarize(mn_area = mean(area))
my_CIACountries

#Base R
my_CIACountries <- CIACountries
my_CIACountries$high_educ <- CIACountries$educ > 4.5
tapply(my_CIACountries$area,my_CIACountries$high_educ, mean)

```


##In class assignment: What first letter has the largest population? 
```{r}
CIACountries %>%
  mutate(fletter = substring(country, 1, 1)) %>%
  group_by(fletter) %>%
  summarise(totalpop = sum(pop)) %>% 
  arrange(-totalpop) %>% 
  head(5)

```
#What proportion of countries produce any oil stratified by net users class? 
```{r}
CIACountries %>% 
  mutate(anyoil = (oil_prod > 0)) %>% 
  group_by(net_users) %>% 
  summarise(oil_prop = mean(anyoil, na.rm = TRUE))
```




## Data viz

```{r}
library(tidyverse)
library(babynames)

#Let's tale about layers!

#scatterplot 
ggplot(aes(x = area, y = pop),data = CIACountries) + geom_point() + geom_smooth(method = "lm")

#histogram
ggplot(aes(x = pop),data = CIACountries) + geom_histogram()

#histogram
ggplot(aes(x = net_users),data = CIACountries) + geom_bar()

#boxplot
ggplot(aes(y = pop),data = CIACountries) + geom_boxplot()

#Kernel Density estimate
ggplot(aes(x = pop),data = CIACountries) + geom_density()

#Add multiple layers!
ggplot(aes(x = area, y = pop),data = CIACountries) + geom_point() + geom_smooth()
ggplot(aes(x = pop),data = CIACountries) + geom_density(color = "red") + geom_histogram(aes(y = ..density..)) 


#Using color
ggplot(aes(x = area, y = pop, color = (oil_prod > 0)),data = CIACountries) + geom_point()

#faceting
ggplot(aes(x = area, y = pop, color = (oil_prod > 0)),data = CIACountries) + geom_point() + facet_wrap( ~ net_users)
```


## Presidential Clustering
```{r eval = FALSE}
library(RCurl)
library(XML)
library(tm)

#Now get them all
doc<-htmlParse(getURL("http://www.presidentialrhetoric.com/historicspeeches/index.html"))
vec<-xpathSApply(doc, "//a/@href")
vecList<-vec[grep("inaugural",vec)]

presList<-list()
for (vvv in vecList){print(vvv)
                     url<-paste0("http://www.presidentialrhetoric.com/historicspeeches/",vvv)
                     a<-getURL(url)
                     b<-htmlParse(url)
                     x<-readHTMLTable(b)
                     text<-x[[6]]$V1
                     #text<-levels(x[[6]]$V1)
                     text<-gsub("\n","",text)
                     text<-gsub("\r","",text)
                     text<-gsub("[,.]","",text)
                     presList[[vvv]]<-PlainTextDocument(text)
}

presCorpus <- Corpus(VectorSource(presList))
#presCorpus<-tm_map(presCorpus,tolower)
presCorpus <- tm_map(presCorpus, content_transformer(tolower))
presCorpus<-tm_map(presCorpus,removeWords,stopwords("english"))
presCorpus<-tm_map(presCorpus,removePunctuation)
presCorpus<-tm_map(presCorpus,removeNumbers)
presCorpus<-tm_map(presCorpus,stripWhitespace)
#presCorpus<-tm_map(presCorpus,stemDocument)
presTDM <-TermDocumentMatrix(presCorpus)
presTDM$dimnames$Docs<-substring(vecList,1,5)
#save(presTDM,file="/Users/gregorymatthews/Dropbox/presTDM.RData")
```

```{r}
library(RCurl)
library(XML)
library(tm)
load("/Users/gregorymatthews/Dropbox/presTDM.RData")
#presTDM$dimnames$Docs <- vecList
#findFreqTerms(presTDM,300)
#findAssocs(presTDM,"drugs",0.9)

m <- as.matrix(t(presTDM))
rownames(m)<-presTDM$dimnames$Docs
library(proxy)
#What is the "cosine" method here?  Why am I using it?
#http://en.wikipedia.org/wiki/Cosine_similarity
#Wikipedia answer everything.  
d <- dist((m),method="cosine")
#d <- dist((m),method="euclidean")
hc <- hclust(d, method="average")
plot(hc)
```

```{r eval = FALSE}
weighted<-weightTfIdf(presTDM)
inspect(weighted[1:10,1:10])
m <- as.matrix(weighted)
rownames(m) <- 1:nrow(m)
cl <- kmeans(t(m), 10)
cl
table(cl$cluster)

#Some other fun stuff
m <- as.matrix(t(presTDM))
rownames(m)<-presTDM$dimnames$Docs
group<-factor(c(rep(1,31),rep(2,24)))
colnames(m)[1]

t.test(m[,1]~group)$p.value

goat<-function(x){
  if (var(x)>0.001){
  out<-t.test(x~group)$p.value
  return(out) } else {
    return(NA)}
}
results<-apply(m,2,goat)
interesting<-data.frame(word=colnames(m),pval=results)
interesting[order(interesting$pval),][1:25,]

((interesting[order(interesting$pval),])[1:25,])[-c(1,4,10),]
t.test((m[,colnames(m)=="duties"])~group)
t.test((m[,colnames(m)=="world"])~group)
t.test((m[,colnames(m)=="public"])~group)
t.test((m[,colnames(m)=="god"])~group)
t.test((m[,colnames(m)=="poverty"])~group)
t.test((m[,colnames(m)=="women"])~group)
t.test((m[,colnames(m)=="men"])~group)
t.test((m[,colnames(m)=="drugs"])~group)
t.test((m[,colnames(m)=="constitution"])~group)
t.test((m[,colnames(m)=="debt"])~group)

t.test((m[,colnames(m)=="streets"])~group)
t.test((m[,colnames(m)=="corruption"])~group)
t.test((m[,colnames(m)=="tyranny"])~group)
t.test((m[,colnames(m)=="faith"])~group)
t.test((m[,colnames(m)=="religion"])~group)
t.test((m[,colnames(m)=="threat"])~group)
t.test((m[,colnames(m)=="blood"])~group)

word <- "children"
counts <- c()
for (i in 1:length(presCorpus)){
  counts[i] <- 0
  if (gregexpr(word,presCorpus[[i]]$content)[[1]]>0){
counts[i] <- length(gregexpr(word,presCorpus[[i]]$content)[[1]])
}
}
  
plot(seq(1777,2005,4)[-c(1:3)],counts)
points(seq(1777,2005,4)[-c(1:3)], counts, type = "l")




m[,colnames(m)=="work"]

temp<-strsplit(presList[[15]]," ")
cbind(temp[[1]][which(temp[[1]]=="duty")+1],
      temp[[1]][which(temp[[1]]=="duty")+2],
      temp[[1]][which(temp[[1]]=="duty")+3])
thing<-cbind(temp[[1]][which(temp[[1]]=="duty")-3],
             temp[[1]][which(temp[[1]]=="duty")-2],
             temp[[1]][which(temp[[1]]=="duty")-1])



#Examples from here: http://michael.hahsler.net/SMU/7337/install/tm.R
```